---
title: "Reproducible Report Template"
output: html_notebook
--- 


The purpose of this template is to recap some of the concepts covered in the course and also to provide a resource of code snippets that will be useful for your course project (and hopefully other future R projects!)

## How to get un-stuck

As you'll prepare this report, you'll likely encounter strange errors and get stuck multiple times. **That's normal and there's no way around it as you learn**. The basic approach for getting un-stuck is to spend a *limited* amount of time *actively* looking for the solution and then *asking for help*. 

Two common scenarios, and a systematic way to troubleshoot are:

1. You encounter an unexpected result or error
  1. Check "spelling" - missing or extra quotes, commas, parentheses; misspelled function names
  2. Read the error message and try to make sense of it
  3. Google the error message (cut and paste!)
2. You know what you want to do but don't know how to do it in R
  1. Consult a relevant RStudio Cheat Sheet (https://www.rstudio.com/resources/cheatsheets/)
  2. Try Google (StackOverflow pages often have helpful info)
  3. Do a Datacamp tutorial (great for a refresher!)

Asking good questions when you get stuck is an art in an of itself. I recommend the following formula:

>  I tried to accomplish (goal) so I tried this code: 
>  (copy and paste the code here)
>  
>  I expected this: ...
>  
>  Instead I get this result / error: 
>  (copy and paste the result or error message)



Below you'll see the *setup* chunk. Run it (by clicking the green triangle) to load the *tidyverse* and *lubridate* packages both of which you'll likely use in your report.

Remember that adding the `include=FALSE` *chunk option* will prevent this code chunk from showing up in your rendered document (i.e. what you'll see when you click *Preview*) if that's what you want.

```{r setup}
library(tidyverse)
library(lubridate)
```


## Background and Objectives

This section should address the following questions:

* What is the question to be addressed by the project?
* How could addressing this question change patient management and/or further generalizable knowledge?
  * What's the point of doing this analysis? 
  * Would anyone else really care about the results?
* What are the first 1-2 key graphs or statistics that you'd like to generate?


## Data Acquisition

This section should contain the following:

* General properties of the data set
* Code Book (see below)
* For data files:
  * Origin, with references
  * File name and file format

The code in this section should accomplish the following tasks:

1. Import data file
2. Isolate data of interest

### Import data file

#### CSV file

The following sample code imports the CSV file named `my_data.csv` into the data frame `my_data`. Remember that the assignment operator `<-` plugs the output of the function on the right into the object on the left. 

```{r}
my_data <- read_csv("my_data.csv")
```

#### Excel file

```{r}
library(readxl)
my_data <- read_excel("my_data.xlsx")
```

### Isolate data of interest

Look at the columns of your dataframe. Which ones do you need for your analysis? Do you want to analyze a subset of your data, e.g. only the data from the last year?

Remember that *tidy functions* always:

1. Take a data frame as their first argument; and
2. Return a data frame.

These two properties let you link together tidy functions in a pipeline with the *pipe operator* `%>%`.

`select()` retains columns by name and `filter()` retains rows by a logical condition. 

The following sample code would select the `Patient_Age`, `Patient_Sex`, and `Date_of_Collection` columns and then filter the rows such that only those rows remain in which `Date_of_Collection` is within 2017. The `arrange()` function would then sort rows by `Date_of_Collection`.

```{r query}
my_data <- my_data %>%
  select(Patient_Age,
         Patient_Sex,
         Date_of_Collection) %>%
  filter(year(Date_of_Collection) == "2017") %>%
  arrange(Date_of_Collection)
```

### Code book

The code book is where you document concisely document the data contained in each column. Be sure to include enough detail to avoid misunderstandings. For example:

- MRN. The patient medical record number at HUP.
- AGE. Patient age at the time the specimen was collected.


## Data Exploration and Cleaning

* Convert messy data to tidy
    * Each variable in its own column
    * Each case in its own row
    * Each value in its own cell
* Check for missing values
* Then, go column by column and do the following:
    * Verify (and if necessary, convert) data type
        * Numeric variables should be integer or numeric
        * Date-time variables should be Date or POSIXct
        * Categorial variables should be factor
        * All others, in general, should be character
    * Visualize distribution
        * Numeric variables: histogram
        * Categorical variables: bar graph or numerical summary
        * Character variables: inspect a small sample
    * Comment on whether the data is as expected
        * Range and shape of distribution
        * Number and breakdown of categories
        * Properties of character variables
    * Explain discrepancies
    * Resolve data problems

#### Missing values

The following code chunk shows the number of missing values for each column. 

```{r}
my_data %>%
    map_df(function(x) sum(is.na(x))) %>%
    gather(column, NAs)
```

#### Numeric variables

`variable_1` is an **integer** column, which is appropriate.

```{r}
ggplot(data = my_data) + 
    geom_histogram(mapping = aes(x = variable_1),
                   binwidth = 1)
```

#### Character variables

`variable_3` is a **character** column, which is appropriate.

```{r}
set.seed(1)
my_data %>%
    sample_n(10) %>%
    pull(variable_3)
```

#### Convert from character to factor

`variable_4` is a **character** column, but since it represents a categorial variable, convert it to a **factor**.

```{r}
my_data <- my_data %>%
    mutate(variable_4 = as_factor(variable_4))

# code to create a data summary
```

#### Convert from character to integer

`variable_5` is a **character** column, but since it represents a numeric variable, convert it to **integer**.

Before doing so, examine any rows that do *not* contain an integer value:

```{r}
my_data %>%
    filter(is.na(as.integer(variable_5))) %>%
    group_by(variable_5) %>%
    summarize(n = n()) %>% 
    arrange(desc(n))
```

Now convert to integer and drop all rows in which `variable_5` had a non-numeric entry.

```{r}
my_data <- my_data %>%
    mutate(variable_5 = as.integer(variable_5)) %>%
    drop_na(variable_5)
```

## Visualization and Modeling

This section will be specific for your analysis. 

The basic template for creating a plot is 

```{r}
ggplot(data = my_data) +
  geom_function(mapping = aes(mappings))
```

- where `my_data` is a tidy data frame,
- geom_function is a "geom" function (like geom_histogram()), and
- mappings map columns of the data frame to features on the graph (like x, y, or color)

You might want to complete the DataCamp course "Intro to the Tidyverse" to review ggplot and keep the ggplot cheat sheet on hand!

## Summary and Conclusions

This is a good place to recap each important result and draw larger conclusions.